Formal Proof of the Relative Consistency of ZF$+\neg$AC with Isabelle/ZF

### 2025-01-08 paper/main.tex
updated

\usepackage{mdframed}
% isabelleコードの枠
\newenvironment{isaframe}{\begin{mdframed}[topline=false, rightline=false, bottomline=false]}{\end{mdframed}}

@@

% 参考文献
\usepackage[backend=biber]{biblatex}
\addbibresource{main.bib}

@@

@@

\chapter*{Abstract}
This paper presents a formalization of the relative consistency of the negation of the Axiom of Choice ($\neg$AC) 
with respect to Zermelo-Fraenkel set theory (ZF) using the proof assistant Isabelle/ZF. 
We use the countable transitive model (c.t.m.) approach, assuming the existence of a c.t.m.\ of ZF, 
and apply forcing to construct a symmetric extension called the basic Cohen model, 
which serves as a model of ZF + $\neg$AC. This work builds on previous formalizations, 
such as Gunther et al.'s work on forcing, and provides the first formalization of the relative consistency of $\neg$AC with ZF.

The proof is based on Karagila's lecture note and Jech's textbook, which align well with the c.t.m.\ approach. 
However, some parts required adopting alternative proof methods.
The formalization process revealed insights into the challenges of formalizing set theory and the potential for future improvements.

@@

Proof assistants are software tools designed to formally verify mathematical proofs, programs, and more. 
These systems are widely used in various fields, including mathematics \cite{four_color,kepler}, cryptography \cite{EasyCrypt_AWS,EasyCrypt_zero_knowledge}, 
hardware verification \cite{Coq_hardware_verification,Coq_hardware_verification_2},
and program verification \cite{survey_proof_assistants_program_verification}. 
Operating on the basis of formal logic, they enable users to interactively construct proofs, which are then checked by the system.
Well-known examples of proof assistants include Isabelle \cite{isabelle_toppage}, Coq \cite{coq_toppage}, Lean \cite{lean_toppage} Agda \cite{agda_toppage}, and HOL Light \cite{hol_light_toppage},
each offering unique features and strengths \cite{survey_theorem_provers}.

@@

In particular, these systems are effective in verifying complex proofs.
For example, the Four Color Theorem \cite{four_color} and Kepler's Conjecture \cite{kepler} are well-known cases of theorems formally proven using proof assistants. 
These large-scale proofs were difficult for humans to verify manually, but their correctness was assured through formal verification.
Similarly, in systems where safety is critical, formal verification of programs can significantly enhance their reliability. 
Notable examples of formally verified systems include the CompCert C compiler \cite{compcert} and the seL4 microkernel \cite{seL4}.

Proof assistants are not limited to applications requiring complex proofs; 
they are also widely used in efforts to formalize various mathematical theorems. 
Fundamental mathematical results, such as those in group and ring theory or calculus, 
have been formalized by many proof assistants, and the scope of such formalizations continues to expand.

@@

While this paper provides a brief explanation of axiomatic set theory, 
a detailed exposition can be found in standard textbooks such as \cite{kunen2011,jech_set_theory}.

@@

Forcing is a technique first introduced by Cohen \cite{cohen} to prove the independence of AC in ZF 
and the independence of CH in ZFC. 
It is a powerful method for exploring models of set theory, 
and it has been further refined by other researchers over time, 
leading to many significant results being established through its use. 
Forcing continues to play a crucial role in contemporary research in axiomatic set theory.

@@

Paulson et al. formalized an extensive part of ZF set theory,
including cardinal arithmetic \cite{paulson_cardinal_AC}, relativization \cite{paulson_AC_consistency}, the reflection theorem \cite{paulson_reflection}, 
features for handling inductive definitions \cite{paulson_datatype_impl}, and the relative consistency of AC with ZF \cite{paulson_AC_consistency}
The proof of the relative consistency was achieved by constructing Gödel's constructible universe.

Building on these results, Gunther et al. formalized forcing and a proof of the independence of CH \cite{gunther_forcing,gunther_independence} in Isabelle/ZF.
In these formalizations, they used the countable transitive model (c.t.m.) approach, 
which we also adopt in this paper.

In Lean, Han and van Doorn also formalized forcing and the independence of CH \cite{flypitch} in Lean 3.
using the Boolean-valued model approach, which is another approach to forcing.
Additionally, in Lean 4, Holmes and Wilshaw formalized the complex parts of the consistency proof of Quine's New Foundations \cite{NF_consistency}, ensuring the correctness of the proof.

@@

The primary reason we chose Isabelle/ZF for this formalization is that Isabelle/ZF is a mature proof assistant for ZF set theory,
in particular, the formalization by \cite{gunther_forcing} is a major advantage for this study.
Although there is also a formalization of forcing in Lean 3 by Han and van Doorn \cite{flypitch}, development for Lean 3 has already ended.

@@

Our proof is based on Karagila's lecture note \cite{karagila} and Jech's book \cite{jech_AC},
as these resources align well with the c.t.m.\  approach.
However, some parts of the proof required modification to fit the formalization in Isabelle/ZF.

@@

\section{Set-Theoretic Preliminaries}
In this section, we introduce the concepts of set theory used in the formalization of this study.
We discuss in ZFC, within which we explore ZF itself. 
Basically, our definition and theorems combine \cite{kunen2011}, \cite{karagila}, and \cite{jech_AC}
adapted to the formalized form in Isabelle/ZF.

@@

\subsection{The c.t.m.\  Approach}\label{subsection:ctm}
This section explains the c.t.m.\ approach.
This approach is a well-established method for proving relative consistency in axiomatic set theory,
as it is covered in \cite{kunen2011}.

@@

First, we introduce some concepts,
following Jech \cite{jech_set_theory}, chapter 6 and 12.

@@

\subsection{Isabelle and Isabelle/ZF}
Isabelle is a proof assistant developed \cite{paulson_1986}, used for formalizing mathematical proofs and verifying programs. 
Since its release in 1986, it has continued to be developed by many researchers.

One notable example of Isabelle's achievements is the formal verification of the seL4 microkernel \cite{seL4}.
This was a massive project involving over a million lines of Isabelle code. 
More recently, in The ALEXANDRIA Project \cite{alexiandria}, various advanced mathematical results have been formalized in Isabelle.

@@

\subsubsection{Recursive Data Types and Functions}
\cite{paulson_datatype_impl} introduced \texttt{"datatype"} to recursive data types within ZF set theory, 
and "primrec" to define functions on top of these data types. 
This made it possible to define data structures such as lists within ZF set theory.
For example, the set of all lists of elements from a set $A$ is defined in a library of Isabelle/ZF as follows:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
consts
    list       :: "i⇒i"

@@

\subsubsection{Internalized Formula and Satisfaction Relation}
Paluson defined formulas encoded as sets in ZF set theory in their work \cite{paulson_AC_consistency},
using \texttt{"datatype"} as follows:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
consts   formula :: i
datatype
    "formula" = Member ("x ∈ nat", "y ∈ nat")
                      | Equal  ("x ∈ nat", "y ∈ nat")
                      | Nand ("p ∈ formula", "q ∈ formula")
                      | Forall ("p ∈ formula")
\end{minted}
\end{isaframe}
Note that \texttt{"nat"} is the set of natural numbers.
The unique logical connective in this definition is NAND and 
other logical connectives such as AND, OR, and NOT on formulas are defined in terms of NAND.
Formulas are represented using the de Brujin index, in other words, 
a occurence of a natural number $n$ in a formula represents the $(n - m)$-th free variable when this $n$ occurs under $m$ quantifiers.

@@

Paulson and Gunther et al. declared locales to handle internal models of ZF set theory in their work \cite{paulson_AC_consistency, gunther_forcing}, respectively.
For example, the locale \texttt{"M\_ctm"} by Gunther et al. fixes \texttt{"M"} and \texttt{"enum"},
and introduces assumptions that \texttt{"M"} holds the axioms of ZF set theory, 
\texttt{"M"} is a transitive set, and \texttt{"enum"} is a enumeration of \texttt{"M"}.
Therefore, we can work in the context of a c.t.m.\  of ZF by working within this locale.

\subsubsection{Forcing}
Gunther et al. formalized forcing method in ZF set theory \cite{gunther_forcing}.
Their formalization follows Kunen's book \cite{kunen2011}, which uses the c.t.m.\  approach.

@@

\chapter{Formalization of the Proof} 
In this chapter, we present the formalization of the relative consistency proof of $\neg$AC with ZF in Isabelle/ZF.
We follow the outline presented in \cref{sec:outline}.
We choose \emph{ZF-Constructible} by \cite{paulson_AC_consistency} as Isabelle theory 
\footnote{Note that this does not mean assuming $V=L$.}
, which includes the ZF axioms along with useful definitions and lemmas built on top of them.
We also reuse the formalization of forcing by \cite{gunther_forcing}.

@@

\subsubsection{$\mathbb{P}$-names}
We begin our work within the locale \texttt{"forcing\_data"} by \cite{gunther_forcing}, 
in which a c.t.m.\  \texttt{"M"} and a notion of forcing \texttt{"P"} are fixed.
The relation $\leq_{\mathbb{P}}$ is denoted by \texttt{"leq"} and 
a maximum element $1_{\mathbb{P}}$ is denoted by \texttt{"one"}.

@@

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition HP_set_succ :: "[i, i] ⇒ i" where 
    "HP_set_succ(a, X) ≡ Pow(X × P) ∩ M"

@@

definition P_names :: "i" where "P_names ≡ { x ∈ M . ∃ a. Ord(a) ∧ x ∈ P_set(a) }" 
\end{minted}
\end{isaframe}

@@

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
locale forcing_data_partial = forcing_data + 
    assumes leq_relation_on_P : "leq ∈ Pow(P × P)" 
    and leq_partial_order : "partial_order_on(P, leq)" 
\end{minted}
\end{isaframe}

We define the set of automorphisms on $\mathbb{P}$, denoted by \texttt{"P\_auto"}, as follows:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition is_P_auto :: "i ⇒ o" where
    "is_P_auto(π) ≡ π ∈ M ∧ π ∈ bij(P, P) ∧ (∀ p ∈ P. ∀q ∈ P. p ≼ q ⟷ π`p ≼ π`q)"

definition P_auto where "P_auto ≡ { π ∈ P → P. is_P_auto(π) }" 
\end{minted}
\end{isaframe}

Where \texttt{"bij(A, B)"} is the set of bijections from \texttt{A} to \texttt{B}
and \texttt{"f`x"} denotes the function application of \texttt{f} to \texttt{x}.
We denote the induced automorphism on $\mathbb{P}$-names by \texttt{π} as \texttt{"Pn\_auto(π)"}.
Our definition of \texttt{"Pn\_auto"} satisfies the equality given in \cref{dfn:automorphism}.

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma Pn_auto : 
    "x ∈ P_names ⟹ Pn_auto(π)`x = { <Pn_auto(π)`y, π`p> . <y, p> ∈ x}"  
\end{minted}
\end{isaframe}

\subsubsection{Groups of Automorphisms and Normal Filters}
First, we define the set of all subgroups of a group of automorphisms \texttt{G} that are in \texttt{"M"}.
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
  definition is_P_auto_group where 
      "is_P_auto_group(G) ≡ 
        G ⊆ { π ∈ P → P. is_P_auto(π) } 
      ∧ (∀π ∈ G. ∀τ ∈ G. π O τ ∈ G) 
      ∧ (∀π ∈ G. converse(π) ∈ G)"

  definition P_auto_subgroups where 
      "P_auto_subgroups(G) ≡ { H ∈ Pow(G) ∩ M. is_P_auto_group(H) }" 
\end{minted}
\end{isaframe}
Where \texttt{"π O τ"} denotes the composition of \texttt{π} and \texttt{τ}.
Then, we declare a locale \texttt{"M\_symmetric\_system"} that fixes a group of automorphisms \texttt{"𝒢"} and a normal filter \texttt{"ℱ"} on \texttt{"𝒢"}.
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
locale M_symmetric_system = forcing_data_partial + 
    fixes 𝒢 ℱ 
    assumes 𝒢_in_M : "𝒢 ∈ M"  
    and 𝒢_P_auto_group : "is_P_auto_group(𝒢)"   
    and ℱ_in_M : "ℱ ∈ M"
    and ℱ_subset : "ℱ ⊆ P_auto_subgroups(𝒢)" 
    and ℱ_nonempty : "ℱ ≠ 0" 
    and ℱ_closed_under_intersection : "∀A ∈ ℱ. ∀B ∈ ℱ. A ∩ B ∈ ℱ" 
    and ℱ_closed_under_supergroup : 
        "∀A ∈ ℱ. ∀B ∈ P_auto_subgroups(𝒢). A ⊆ B ⟶ B ∈ ℱ" 
    and ℱ_normal : "∀H ∈ ℱ. ∀π ∈ 𝒢. { π O τ O converse(π). τ ∈ H } ∈ ℱ" 
\end{minted}
\end{isaframe}

@@

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition sym where "sym(x) ≡ { π ∈ 𝒢. Pn_auto(π)`x = x }"

@@

definition HS where "HS ≡ { x ∈ P_names. ∃a. Ord(a) ∧ x ∈ HS_set(a) }" 
\end{minted}
\end{isaframe}

@@

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma HS_iff: "x ∈ HS ⟷ x ∈ P_names ∧ domain(x) ⊆ HS ∧ symmetric(x)" 
\end{minted}
\end{isaframe}

@@

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition SymExt where "SymExt(G) ≡ { val(G, x). x ∈ HS }" 
\end{minted}
\end{isaframe}

Where \texttt{"val(G, x)"} is the interpretation $x^G$ formalized by \cite{gunther_forcing}.

We declare a locale \texttt{"M\_symmetric\_system\_G\_generic"},
which combines the locale \texttt{"M\_symmetric\_system"} and the locale \texttt{"G\_generic"}.
\texttt{"G\_generic"} is a locale, defined by \cite{gunther_forcing}, fixes a generic filter \texttt{"G"} on \texttt{"P"}.

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
locale M_symmetric_system_G_generic = M_symmetric_system + G_generic
\end{minted}
\end{isaframe}

@@

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma M_subset_SymExt : "M ⊆ SymExt(G)" 
lemma SymExt_subset_GenExt : "SymExt(G) ⊆ GenExt(G)" 
lemma Transset_SymExt : "Transset(SymExt(G))" 
lemma SymExt_countable : "nat ≈ SymExt(G)" 
\end{minted}
\end{isaframe}

@@

We utilized the automated formula synethisis method by \cite{gunther_forcing};
however, it could not applied to complex formulas, 
and re-verifying the properties of the derived formulas in a more usable form remained a repititive task.

@@

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma sats_is_P_name_fm_iff : 
    fixes env i j x 
    assumes "env ∈ list(M)" "i < length(env)" "j < length(env)" 
                    "nth(i, env) = P" "nth(j, env) = x"  
    shows "sats(M, is_P_name_fm(i, j), env) ⟷ x ∈ P_names"

@@

lemma sats_is_HS_fm_iff : 
    fixes x i j env 
    assumes "env ∈ list(M)" "i < length(env)" "j < length(env)" 
                    "nth(i, env) = <ℱ, 𝒢, P, P_auto>" "nth(j, env) = x" 
    shows "sats(M, is_HS_fm(i, j), env) ⟷ x ∈ HS" 
\end{minted}
\end{isaframe}

@@

In this section, we prove that symmetric extensions are models of ZF.
Initially, we attempted to formalize this based on its proof in \cite[Theorem 10.10]{karagila},
but we found that additional work was required to formalize parts that are omitted in the note.
Therefore, we decided to prove this by using the relativized forcing relation $\Vdash_{\mathrm{HS}}$.
Initial approach is a traditional method for proving that symmetric extensions are models of ZF,
and applicable to a more general situation since 
it does not rely on forcing or generic and symmetric extensions.
whereas our approach is more specific to the context of symmetric extensions.

>>>>

\usepackage{mdframed}

@@

% 参考文献
\usepackage[backend=biber,style=authoryear]{biblatex}
\addbibresource{main.bib}

@@

% isabelleコードの枠
\newenvironment{isaframe}
  {\begin{mdframed}[topline=false, rightline=false, bottomline=false]}
  {\end{mdframed}}
\newenvironment{isaframe2}[2]
  {\begin{mdframed}[topline=false, rightline=false, bottomline=false, frametitle={{\tiny \href{#1}{#2}}}]}
  {\end{mdframed}}

@@

\chapter*{Abstract}
This paper presents a formalization of the relative consistency of the negation of the Axiom of Choice ($\neg$AC) 
with respect to Zermelo-Fraenkel set theory (ZF) using the proof assistant Isabelle/ZF. 
We use the countable transitive model (c.t.m.) approach, assuming the existence of a c.t.m.\ of ZF, 
and apply forcing to construct a symmetric extension called the basic Cohen model, 
which serves as a model of ZF + $\neg$AC. This work builds on previous formalizations, 
such as forcing (Gunther et al. 2020), and provides the first formalization of the relative consistency of $\neg$AC with ZF.
The proof is based on (Karagila 2023) and (Jech 2008), which align well with the c.t.m.\ approach,
although the proof that the symmetric extension satisfies the axioms of ZF required modification to fit this formalization.
The formalization process gave insights into the challenges of formalizing set theory and the potential for future improvements.

@@

Proof assistants are software tools designed to formally verify mathematical proofs, programs, and more. 
These systems are widely used in various fields, including mathematics \parencite{four_color,kepler}, cryptography \parencite{EasyCrypt_AWS,EasyCrypt_zero_knowledge}, 
hardware verification \parencite{Coq_hardware_verification,Coq_hardware_verification_2},
and program verification \parencite{survey_proof_assistants_program_verification}. 
Operating on the basis of formal logic, they enable users to interactively construct proofs, which are then checked by the system.
Well-known examples of proof assistants include Isabelle \parencite{isabelle_toppage}, Coq \parencite{coq_toppage}, Lean \parencite{lean_toppage} Agda \parencite{agda_toppage}, and HOL Light \parencite{hol_light_toppage},
each offering unique features and strengths \parencite{survey_theorem_provers}.

@@

In particular, these systems are effective in verifying proofs involving intricate case analysis.
For example, the Four Color Theorem \parencite{four_color} and Kepler's Conjecture \parencite{kepler} are well-known cases of theorems formally proven using proof assistants. 
These large-scale proofs were difficult for humans to verify manually, but their correctness was assured through formal verification.
Similarly, in systems where safety is critical, formal verification of programs can significantly enhance their reliability. 
Notable examples of formally verified systems include the CompCert C compiler \parencite{compcert} and the seL4 microkernel \parencite{seL4}.

Proof assistants are not limited to applications requiring complex case analysis;
they are also widely used in efforts to formalize various mathematical theorems. 
Fundamental mathematical results, such as those in group and ring theory or calculus, 
have been formalized by many proof assistants, and the scope of such formalizations continues to expand.
For example, mathlib \parencite{mathlib} is a library actively developed in Lean, including formalizations of various mathematical concepts and theorems.

@@

While this paper provides a brief explanation of axiomatic set theory, 
a detailed exposition can be found in standard textbooks such as \parencite{kunen2011,jech_set_theory}.

@@

Forcing is a technique first introduced by Cohen \parencite{cohen} to prove the independence of AC in ZF 
and the independence of CH in ZFC. 
It is a powerful method for exploring models of set theory, 
and it has been further refined by other researchers over time, 
leading to many significant results being established through its use. 
Forcing continues to play a crucial role in contemporary research in axiomatic set theory.

@@

Paulson et al. formalized an extensive part of ZF set theory,
including cardinal arithmetic \parencite{paulson_cardinal_AC}, relativization \parencite{paulson_AC_consistency}, the reflection theorem \parencite{paulson_reflection}, 
features for handling inductive definitions \parencite{paulson_datatype_impl}, and the relative consistency of AC with ZF \parencite{paulson_AC_consistency}
The proof of the relative consistency was achieved by constructing Gödel's constructible universe.

Building on these results, Gunther et al. formalized forcing and a proof of the independence of CH \parencite{gunther_forcing,gunther_independence} in Isabelle/ZF.
In these formalizations, they used the countable transitive model (c.t.m.) approach, 
which we also adopt in this paper.

In Lean, Han and van Doorn also formalized forcing and the independence of CH \parencite{flypitch} in Lean 3.
using the Boolean-valued model approach, which is another approach to forcing.
Additionally, in Lean 4, Holmes and Wilshaw formalized the complex parts of the consistency proof of Quine's New Foundations \parencite{NF_consistency}, ensuring the correctness of the proof.

@@

The primary reason we chose Isabelle/ZF for this formalization is that Isabelle/ZF is a mature proof assistant for ZF set theory,
in particular, the formalization by \parencite{gunther_forcing} is a major advantage for this study.
Although there is also a formalization of forcing in Lean 3 by Han and van Doorn \parencite{flypitch}, development for Lean 3 has already ended.

@@

Our proof is based on Karagila's lecture note \parencite{karagila} and Jech's book \parencite{jech_AC},
as these resources align well with the c.t.m.\  approach.
However, some parts of the proof required modification to fit the formalization in Isabelle/ZF.

@@

\section{Set-Theoretic Preliminaries}
In this section, we introduce the concepts of set theory used in the formalization of this study.
We discuss in ZFC, within which we explore ZF itself. 
Basically, our definition and theorems combine \parencite{kunen2011}, \parencite{karagila}, and \parencite{jech_AC}
adapted to the formalized form in Isabelle/ZF.

@@

\subsection{The c.t.m.\  Approach}\label{subsection:ctm}
This section explains the c.t.m.\ approach.
This approach is a well-established method for proving relative consistency in axiomatic set theory,
as it is covered in \parencite{kunen2011}.

@@

First, we introduce some concepts,
following Jech \parencite{jech_set_theory}, chapter 6 and 12.

@@

\subsection{Isabelle and Isabelle/ZF}
Isabelle is a proof assistant developed \parencite{paulson_1986}, used for formalizing mathematical proofs and verifying programs. 
Since its release in 1986, it has continued to be developed by many researchers.

One notable example of Isabelle's achievements is the formal verification of the seL4 microkernel \parencite{seL4}.
This was a massive project involving over a million lines of Isabelle code. 
More recently, in The ALEXANDRIA Project \parencite{alexiandria}, various advanced mathematical results have been formalized in Isabelle.

@@

\subsubsection{Recursive Data Types and Functions}
\parencite{paulson_datatype_impl} introduced \texttt{"datatype"} to recursive data types within ZF set theory, 
and "primrec" to define functions on top of these data types. 
This made it possible to define data structures such as lists within ZF set theory.
For example, the set of all lists of elements from a set $A$ is defined in a library of Isabelle/ZF as follows:
\begin{isaframe} 
\begin{minted}[breaklines]{isabelle}
consts
    list       :: "i⇒i"

@@

\subsubsection{Internalized Formula and Satisfaction Relation}
Paluson defined formulas encoded as sets in ZF set theory in their work \parencite{paulson_AC_consistency},
using \texttt{"datatype"} as follows:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
consts   formula :: i
datatype
    "formula" = Member ("x ∈ nat", "y ∈ nat")
                      | Equal  ("x ∈ nat", "y ∈ nat")
                      | Nand ("p ∈ formula", "q ∈ formula")
                      | Forall ("p ∈ formula")
\end{minted}
\end{isaframe}
Note that \texttt{"nat"} is the set of natural numbers.
The unique logical connective in this definition is NAND and 
other logical connectives such as AND, OR, and NOT on formulas are defined in terms of NAND.
Formulas are represented using the de Brujin index, in other words, 
a occurence of a natural number $n$ in a formula represents the $(n - m)$-th free variable when this $n$ occurs under $m$ quantifiers.

@@

Paulson and Gunther et al. declared locales to handle internal models of ZF set theory in their work \parencite{paulson_AC_consistency, gunther_forcing}, respectively.
For example, the locale \texttt{"M\_ctm"} by Gunther et al. fixes \texttt{"M"} and \texttt{"enum"},
and introduces assumptions that \texttt{"M"} holds the axioms of ZF set theory, 
\texttt{"M"} is a transitive set, and \texttt{"enum"} is a enumeration of \texttt{"M"}.
Therefore, we can work in the context of a c.t.m.\  of ZF by working within this locale.

\subsubsection{Forcing}
Gunther et al. formalized forcing method in ZF set theory \parencite{gunther_forcing}.
Their formalization follows Kunen's book \parencite{kunen2011}, which uses the c.t.m.\  approach.

@@

\chapter{Formalization of the Proof} 
In this chapter, we present the formalization of the relative consistency proof of $\neg$AC with ZF in Isabelle/ZF.
We follow the outline presented in \cref{sec:outline}.
We choose \emph{ZF-Constructible} by \parencite{paulson_AC_consistency} as Isabelle theory 
\footnote{Note that this does not mean assuming $V=L$.}
, which includes the ZF axioms along with useful definitions and lemmas built on top of them.
We also reuse the formalization of forcing by \parencite{gunther_forcing}.

@@

\subsubsection{$\mathbb{P}$-names}
We begin our work within the locale \texttt{"forcing\_data"} by \parencite{gunther_forcing}, 
in which a c.t.m.\  \texttt{"M"} and a notion of forcing \texttt{"P"} are fixed.
The relation $\leq_{\mathbb{P}}$ is denoted by \texttt{"leq"} and 
a maximum element $1_{\mathbb{P}}$ is denoted by \texttt{"one"}.

@@

\begin{isaframe2}
    {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/P_Names.thy\#L10}
    {P\_Names.thy}
\begin{minted}[breaklines]{isabelle}
definition HP_set_succ :: "[i, i] ⇒ i" where 
    "HP_set_succ(a, X) ≡ Pow(X × P) ∩ M"

@@

definition P_names :: "i" where "P_names ≡ { x ∈ M . ∃ a. Ord(a) ∧ x ∈ P_set(a) }" 
\end{minted}
\end{isaframe2}

@@

\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/HS_Definition.thy\#L7}
  {forcing\_data\_partial.thy}
\begin{minted}[breaklines]{isabelle}
locale forcing_data_partial = forcing_data + 
    assumes leq_relation_on_P : "leq ∈ Pow(P × P)" 
    and leq_partial_order : "partial_order_on(P, leq)" 
\end{minted}
\end{isaframe2}

We define the set of automorphisms on $\mathbb{P}$, denoted by \texttt{"P\_auto"}, as follows:
\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/Automorphism_Definition.thy\#L25}
  {Automorphism\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
definition is_P_auto :: "i ⇒ o" where
    "is_P_auto(π) ≡ π ∈ M ∧ π ∈ bij(P, P) ∧ (∀ p ∈ P. ∀q ∈ P. p ≼ q ⟷ π`p ≼ π`q)"

definition P_auto where "P_auto ≡ { π ∈ P → P. is_P_auto(π) }" 
\end{minted}
\end{isaframe2}

Where \texttt{"bij(A, B)"} is the set of bijections from \texttt{A} to \texttt{B}
and \texttt{"f`x"} denotes the function application of \texttt{f} to \texttt{x}.
We denote the induced automorphism on $\mathbb{P}$-names by \texttt{π} as \texttt{"Pn\_auto(π)"}.
Our definition of \texttt{"Pn\_auto"} satisfies the equality given in \cref{dfn:automorphism}.
\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/Automorphism_Definition.thy\#L155}
  {Automorphism\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
lemma Pn_auto : 
    "x ∈ P_names ⟹ Pn_auto(π)`x = { <Pn_auto(π)`y, π`p> . <y, p> ∈ x}"  
\end{minted}
\end{isaframe2}

\subsubsection{Groups of Automorphisms and Normal Filters}
First, we define the set of all subgroups of a group of automorphisms \texttt{G} that are in \texttt{"M"}.
\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/HS_Definition.thy\#L10}
  {HS\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
  definition is_P_auto_group where 
      "is_P_auto_group(G) ≡ 
        G ⊆ { π ∈ P → P. is_P_auto(π) } 
      ∧ (∀π ∈ G. ∀τ ∈ G. π O τ ∈ G) 
      ∧ (∀π ∈ G. converse(π) ∈ G)"

  definition P_auto_subgroups where 
      "P_auto_subgroups(G) ≡ { H ∈ Pow(G) ∩ M. is_P_auto_group(H) }" 
\end{minted}
\end{isaframe2}
Where \texttt{"π O τ"} denotes the composition of \texttt{π} and \texttt{τ}.
Then, we declare a locale \texttt{"M\_symmetric\_system"} that fixes a group of automorphisms \texttt{"𝒢"} and a normal filter \texttt{"ℱ"} on \texttt{"𝒢"}.
\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/SymExt_Definition.thy\#L7}
  {SymExt\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
locale M_symmetric_system = forcing_data_partial + 
    fixes 𝒢 ℱ 
    assumes 𝒢_in_M : "𝒢 ∈ M"  
    and 𝒢_P_auto_group : "is_P_auto_group(𝒢)"   
    and ℱ_in_M : "ℱ ∈ M"
    and ℱ_subset : "ℱ ⊆ P_auto_subgroups(𝒢)" 
    and ℱ_nonempty : "ℱ ≠ 0" 
    and ℱ_closed_under_intersection : "∀A ∈ ℱ. ∀B ∈ ℱ. A ∩ B ∈ ℱ" 
    and ℱ_closed_under_supergroup : 
        "∀A ∈ ℱ. ∀B ∈ P_auto_subgroups(𝒢). A ⊆ B ⟶ B ∈ ℱ" 
    and ℱ_normal : "∀H ∈ ℱ. ∀π ∈ 𝒢. { π O τ O converse(π). τ ∈ H } ∈ ℱ" 
\end{minted}
\end{isaframe2}

@@

\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/HS_Definition.thy\#L32}
  {HS\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
definition sym where "sym(x) ≡ { π ∈ 𝒢. Pn_auto(π)`x = x }"

@@

definition HS where "HS ≡ { x ∈ P_names. ∃a. Ord(a) ∧ x ∈ HS_set(a) }" 
\end{minted}
\end{isaframe2}

@@

\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/HS_Definition.thy\#L90}
  {HS\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
lemma HS_iff: "x ∈ HS ⟷ x ∈ P_names ∧ domain(x) ⊆ HS ∧ symmetric(x)" 
\end{minted}
\end{isaframe2}

@@

\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/SymExt_Definition.thy\#L10}
  {SymExt\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
definition SymExt where "SymExt(G) ≡ { val(G, x). x ∈ HS }" 
\end{minted}
\end{isaframe2}

Where \texttt{"val(G, x)"} is the interpretation $x^G$ formalized by \parencite{gunther_forcing}.

We declare a locale \texttt{"M\_symmetric\_system\_G\_generic"},
which combines the locale \texttt{"M\_symmetric\_system"} and the locale \texttt{"G\_generic"}.
\texttt{"G\_generic"} is a locale, defined by \parencite{gunther_forcing}, fixes a generic filter \texttt{"G"} on \texttt{"P"}.

\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/SymExt_Definition.thy\#L15}
  {SymExt\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
locale M_symmetric_system_G_generic = M_symmetric_system + G_generic
\end{minted}
\end{isaframe2}

@@

\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/SymExt_Definition.thy\#L18}
  {SymExt\_Definition.thy}
\begin{minted}[breaklines]{isabelle}
lemma M_subset_SymExt : "M ⊆ SymExt(G)" 
lemma SymExt_subset_GenExt : "SymExt(G) ⊆ GenExt(G)" 
lemma Transset_SymExt : "Transset(SymExt(G))" 
lemma SymExt_countable : "nat ≈ SymExt(G)" 
\end{minted}
\end{isaframe2}

@@

We utilized the automated formula synethisis method by \parencite{gunther_forcing};
however, it could not applied to complex formulas, 
and re-verifying the properties of the derived formulas in a more usable form remained a repititive task.

@@

\begin{isaframe2}
  {https://github.com/tarakojo/ZF_notAC/blob/177e94cf29db6fb3b0be1e4b2f84d5cd7508a7f4/code/P_Names_M.thy\#L271}
  {P\_Names\_M.thy}
\begin{minted}[breaklines]{isabelle}
lemma sats_is_P_name_fm_iff : 
    fixes env i j x 
    assumes "env ∈ list(M)" "i < length(env)" "j < length(env)" 
                    "nth(i, env) = P" "nth(j, env) = x"  
    shows "sats(M, is_P_name_fm(i, j), env) ⟷ x ∈ P_names"

@@

lemma sats_is_HS_fm_iff : 
    fixes x i j env 
    assumes "env ∈ list(M)" "i < length(env)" "j < length(env)" 
                    "nth(i, env) = <ℱ, 𝒢, P, P_auto>" "nth(j, env) = x" 
    shows "sats(M, is_HS_fm(i, j), env) ⟷ x ∈ HS" 
\end{minted}
\end{isaframe2}

@@

In this section, we prove that symmetric extensions are models of ZF.
Initially, we attempted to formalize this based on its proof in \parencite[Theorem 10.10]{karagila},
but we found that additional work was required to formalize parts that are omitted in the note.
Therefore, we decided to prove this by using the relativized forcing relation $\Vdash_{\mathrm{HS}}$.
Initial approach is a traditional method for proving that symmetric extensions are models of ZF,
and applicable to a more general situation since 
it does not rely on forcing or generic and symmetric extensions.
whereas our approach is more specific to the context of symmetric extensions.
